---
title: 'Steven Maharaj 695281 Assignment 2, Question 3 MAST90125: Bayesian Statistical Learning'
header-includes:
   - \usepackage{bm}
   - \usepackage{amsmath}
output: 
  pdf_document:
    number_sections: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Due: Friday 20 September 2019**  
\vspace{5 mm}

**There are places in this assignment where R code will be required. Therefore set the random seed so assignment is reproducible.**

```{r}
set.seed(695281)#Please change random seed to your student id number.
library(dplyr)
library(tidyr)
library(ggplot2)
library(TruncatedNormal)
```

---
PART A


```{r}
Contraceptionsubset <- read.csv("Contraceptionsubset.csv")
```

---

Part B

Case $x \in (0,\infty)$
```{r}
mu = 0.5
sigma =  1 # sigma

a <- 0
b <- 10 # A number going to infinity
n <- 5000

invtn <- function(n,mu,sigma,a,b){
  u <- runif(n)
  return(qnorm(pnorm(a)+u*(pnorm(b) - pnorm(a)))*sigma + mu)
}

# Using my inverse cdf method
x_empircal <- invtn(n,mu,sigma,a,b)

# using the TruncatedNormal package 
x_empircal <- rtnorm(n = n, mu = mu, lb = a, ub = b, method = "fast")
empircal_den <- density(x_empircal)

x <- seq(a,b,length.out = n)
y <- dnorm(empircal_den$x,mean=mu,sd=sigma)/(pnorm(b,mean=mu,sd=sigma)-pnorm(a,mean=mu,sd=sigma))

plot(empircal_den$x,empircal_den$y,
main="Comparing densities for modified inverse CDF method",
ylab="Density",
xlab = "x",
type="l",
col="blue")
lines(empircal_den$x, y, col="red")
legend("topright",
c("Empirical","True"),
fill=c("blue","red")
)

```



Case $x \in (-\infty,0)$
```{r}
mu = 0.5
sigma =  1 # sigma

a <- -10
b <- 0 # A number going to infinity
n <- 5000


# Using my inverse cdf method
x_empircal <- invtn(n,mu,sigma,a,b)

# using the TruncatedNormal package 
x_empircal <- rtnorm(n = n, mu = mu, lb = a, ub = b, method = "fast")

empircal_den <- density(x_empircal)

x <- seq(a,b,length.out = n)

y <- dnorm(empircal_den$x,mean=mu,sd=sigma)/(pnorm(b,mean=mu,sd=sigma)-pnorm(a,mean=mu,sd=sigma))

plot(empircal_den$x,empircal_den$y,
main="Comparing densities for modified inverse CDF method",
ylab="Density",
xlab = "x",
type="l",
col="blue")
lines(empircal_den$x, y, col="red")
legend("topleft",
c("Empirical","True"),
fill=c("blue","red")
)

```


```{r}
library(TruncatedNormal)
rtnorm(n = 10, mu = 2, lb = 1:10, ub = 2:11, method = "fast")```

```


___
PART C 

```{r}
#Arguments are
#X:  matrix of predictors dimension n times p with flat prior. Includes the intercept. 
#Z:  matrix of predictors dimension n times q with normal prior for u.
#y:  response vector, length p.
#taue_0,tauu_0: initial value for the residual and random effect precision. 
#a.u,b.u. Hyper-parameter for gamma prior for tau_u
#a.e,b.e. Hyper-parameter for gamma prior for tau_e
#iter: number of iterations
#burnin: number of initial iterations to remove
normalmm.Gibbs<-function(iter,Z,X,y,burnin,taue_0,tauu_0,a.u,b.u,a.e,b.e){
  n   <-length(y) #no. observations
  p   <-dim(X)[2] #no of fixed effect predictors.
  q   <-dim(Z)[2] #no of random effect levels
  tauu<-tauu_0
  taue<-taue_0
  #starting value for u.
  u0   <-rnorm(q,0,sd=1/sqrt(tauu))

  #Building combined predictor matrix.
  W<-cbind(X,Z)            #for the joint conditional posterior for b,u
  WTW <-crossprod(W)
  library(mvtnorm)
  
  #storing results.
  par <-matrix(0,iter,p+q+2)  #p beta coefficient, q u coefficients and 2 precision coefficient.
  
  #Create modified identity matrix for joint posterior.
  I0  <-diag(p+q)
  diag(I0)[1:p]<-0
  
  for(i in 1:iter){
    #Conditional posteriors.
    tauu <-rgamma(1,a.u+0.5*q,b.u+0.5*sum(u0^2)) #sample tau_u
    #Updating component of normal posterior for beta,u
    Prec <-WTW + tauu*I0/taue
    P.mean <- solve(Prec)%*%crossprod(W,y)
    P.var  <-solve(Prec)/taue
    betau <-rmvnorm(1,mean=P.mean,sigma=P.var) #sample beta, u
    betau <-as.numeric(betau)
    err   <- y-W%*%betau
    #sample tau_e
    #storing iterations for beta, u, and standard deviation of e, u.
    par[i,]<-c(betau,1/sqrt(tauu),1)
    u0     <-betau[p+1:q]  #extracting u so we can update tau_u.
  }
  
par <-par[-c(1:burnin),] #removing initial iterations
colnames(par)<-c(paste('beta',1:p,sep=''),paste('u',1:q,sep=''),'sigma_b','sigma_e')  
 return(par) 
}
```

