---
title: 'Steven Maharaj 695281 Assignment 2, Question 3 MAST90125: Bayesian Statistical Learning'
header-includes:
   - \usepackage{bm}
   - \usepackage{amsmath}
   - \usepackage{amsfonts}
   - \usepackage{amssymb}
output: 
  pdf_document:
    number_sections: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Due: Friday 20 September 2019**  
\vspace{5 mm}

**There are places in this assignment where R code will be required. Therefore set the random seed so assignment is reproducible.**

```{r}
set.seed(695281)  #Please change random seed to your student id number.
library(dplyr)
library(ggplot2)
library(tidyr)
library(TruncatedNormal)
```


```{r}
rtn <- function(n,b,a,mu,Sigma){
  u <- runif(n)
  g <- pnorm((b-mu)/Sigma) - pnorm((a-mu)/Sigma)
  x <-qnorm((g) * u + pnorm((a-mu)/Sigma))*Sigma + mu 
}

```

___
PART C

We mplement a Gibbs sampler to fit the same mixed model, but now with a probit link.

Assumsing,

 - $p(\bm \beta) \propto 1$
 - $p(\bm u) = \mathcal{N}(\bm 0,\sigma^2_u \bm I)$
 - $p(\tau_u) = Ga(\alpha_u,\gamma_u)$
 
 It can be shown that we have the folling conditional posteriors
 
 $$p\left(\tau_{u} | \cdot\right)=\mathrm{Ga}\left(\alpha_{u}+q / 2, \gamma_{u}+\mathbf{u}^{\prime} \mathbf{u} / 2\right)$$

$$
\left.p\left(\left(\begin{array}{c}{\beta} \\ {u}\end{array}\right) | \cdot\right)=\mathcal{N}\left(\begin{array}{cc}{X^{\prime} X} & {X^{\prime} Z} \\ {Z^{\prime} X} & {Z^{\prime} Z+\tau_{u}  \bm{I}^{-1}}\end{array}\right)^{-1}\left(\begin{array}{c}{X^{\prime} z} \\ {Z^{\prime} z}\end{array}\right),\left(\begin{array}{cc}{X^{\prime} X} & {X'Z} \\ {Z^{\prime} X} & {Z'Z+\tau_{u}  \bm{I}^{-1}}\end{array}\right)^{-1}\right)
$$


We define our inputs for the Gibbs Sampler


```{r}

```

Construct a Gibbs sampler
```{r}

```


